---
title: "03a_ch1_analysis_new"
author: "Kirsten Fuller"
date: "1/18/2022"
output: html_document
---
# Introduction

## Set Working Directory
```{r setup, include=FALSE}
rm( list = ls() ); gc( )

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Boise_State/thesis/ms_thesis_ch1")
```

## Prep Workspace
```{r, include = FALSE, message=FALSE}
# load necessary packages
# define packages
packages <- c("coda", "rjags", "dplyr", "tidyr", "jagsUI", "ggmcmc", "sf") # jagsUI = JAGS user interface

# install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# load packages
invisible(lapply(packages, library, character.only = TRUE))

# set options related to packages
options( dplyr.width = Inf, dplyr.print_min = 100 ) # see all columns and more than 10 rows
```

# Basic analysis steps:
1. Collect and package data
2. Write a model file in BUGS language
3. Set initial values
4. Specify parameters to monitor
5. Set MCMC variables and run analysis
6. Optionally, generate more posterior samples using the update method

## Load Data
```{r}
hooting.df <- read.csv("data/output/ch1_hooting_data")

# we technically had 4 surveys outside of the survey period, let's eliminate those
hooting.df <- hooting.df %>%
  filter(julian_date < 244) %>%
  filter(julian_date > 62)
```

## Define Data
Create separate df's for site level data and survey level data and standardize the predictor values.
```{r}
# SURVEYS:
# define the primary seasons (min to max)
yrs <- min(hooting.df$survey_year):max(hooting.df$survey_year)
yrs  # this includes years when no sampling occurred, that is what we want

# define total length of data period
K <- length(yrs)
K # 21 years


# SITES:
# define the number of sites
I <- length(unique(hooting.df$pac_surveyed))
I # 53 sites

# define the max number of replicate surveys
J <- max(hooting.df$survey_number)
J # 6 is the max number of replicate surveys (probably Bright Angel PAC, which gets surveyed very frequently each year)

# define the number of surveys conducted
T <- nrow(hooting.df)
T # 306 surveys conducted

# add site (pac) number as a new column to the hooting data frame
hooting.df$siteid <- as.numeric(as.factor(hooting.df$pac_surveyed)) 
# JAGS will expect the sites to be numbered

# view the data to make sure those things worked
head(hooting.df)


# DEFINE SITE LEVEL DATA SURVEY LEVEL DATA
# extract site level covariates of interest:
X <- hooting.df %>% # define df  
  dplyr::group_by(siteid) %>% # group by siteid
  slice(1) %>% # select only first row since other values are replicates
  dplyr::select(COLEOGYNE, PJCON, MSHRUB, canyon_geometry, people, supai, redwall, siteid) 

# standardize predictors
X[,1:7] <- apply( X[,1:7], MARGIN = 2, FUN = scale )

# view
head(X); dim(X)

# extract survey level covariates
XJ <- hooting.df %>% 
  mutate(yearid = as.numeric(as.factor(survey_year))) %>%
  dplyr::select(pac_area, julian_date, siteid, yearid, surveyid = survey_number, y_obs = presence)

# standardize covariates of interest
XJ[,1:2] <- apply( XJ[,1:2], MARGIN = 2, FUN = scale )

# view
head(XJ)
```

## Define MCMC Sampling Settings
```{r}
nt <- 5 # thinning = a measure of how much the MCMC chains should be thinned out before storing them. Default is 1. A value of 5 means that we keep every 5th value and discard the others.
nb <- 10000 # burning = the chains we get rid of at the beginning of an mcmc run (10,000 seems high) before it converges; is this the same as iterations?
nc <- 5 # chains = the number of chains, default is 4 I believe
```

Logit(psi)[yr=1] = int + random intercept[yr = 1] 
plot(puts it back on the real scale plogis aka inverse logit(model results$mean$int.psi/ modelresults$mean$eps.K))

Summary of my models:
Model 1 - Full additive model, random effect of year only on detection sub model

Model 2 - Move all of the variables from colonization on to the persistence model, random effect of year only on detection model

Model 3 - Collapse colonization and persistence into the occupancy model and only have an occupancy model and detection model. Add random effect of year to both sub models. 
  
  
## Write Model 1
Dynamic occupancy model
```{r}
sink("dm1.txt") # sink diverts R output to a connection, in this case a text file
cat("model{ 

# PRIORS
########

  # ecological model:    
      # psi = occupancy in year 1
      psi1 ~ dbeta(4, 4) 
# NOTE: using beta instead of binomial here because binomial models the number of successes, but beta models the probability of success. alternative prior from a uniform distribution, dunif(0, 1) = uniform distribution constrained between 0 and 1
      # phi = persistence probability
      int.phi <- log(mean.phi / (1 - mean.phi)) 
      mean.phi ~ dbeta(4, 4)
      # gamma = colonization probability
      int.gam <- log( mean.gam / (1 - mean.gam)) 
      mean.gam ~ dbeta(4, 4)
  # for detection model:
      # p = detection
      int.p <- log( mean.p / (1 - mean.p))
      mean.p ~ dbeta(4, 4)
  # random year intercept for ecological model and for detection model
      for (k in 1:K){ 
      eps.p[ k ] ~ dnorm( 0, prec.p ) T(-7, 7)  # for detection model
      }
      # associated precision of random intercepts:     
      prec.p <- 1 / ( sigma.p * sigma.p )
      sigma.p ~ dt( 0, 2.5, 7 ) T( 0, )
  # priors for beta predictors:
      for( q in 1:2 ){ # loop over number of predictors
        beta.phi[ q ] ~ dnorm( 0, 0.1 ) # prior for persistence predictors
      }
      for( q in 1:3 ){ # loop over number of predictors
        beta.gam[ q ] ~ dnorm( 0, 0.1 ) # prior for colonization predictors
      } # Q
      for( q in 1:2 ){ # loop over number of predictors
        beta.p[ q ] ~ dnorm( 0, 0.1 ) # prior for detection predictors
      } # Q
     
# MODEL
        for( i in 1:I ){  # loop over sites
          # estimating psi[,1]: year 1
          z[ i, 1 ] ~ dbern( psi[ i, 1 ] ) # bernouli dist. of occupancy/non-occupancy
          psi[ i, 1 ] <- psi1
     
          for ( k in 2:K ) { # loop over years
              # occupancy state:
              z[ i, k ] ~ dbern( psi[ i, k ] ) 
              # probability of occupancy starts on year 2
              psi[ i, k ] <- z[ i, k-1 ] * phi[ i, k-1 ] + 
                        ( 1 - z[ i, k-1 ] ) * gam[ i, k-1 ] 
              # model for persistence probability:
              logit( phi[ i, k-1 ] ) <- int.phi + 
                  # fixed predictors:
                  beta.phi[ 1 ] * X[ i, 1 ] + # coleogyne
                  beta.phi[ 2 ] * X[ i, 5 ] # people
              # model for colonization probability:
              logit( gam[ i, k-1 ] ) <- int.gam + 
                  beta.gam[ 1 ] * X[ i, 4 ] + # geometry
                  beta.gam[ 2 ] * X[ i, 3 ] + # mshrub
                  beta.gam[ 3 ] * X[ i, 2 ] # pjcon
          } # K years
        } # I sites
        
      # detection model:
          for( t in 1:T ){  # loop over all replicates (T is the number of rows)
              logit( p[ t ] ) <- int.p +
                  # random effects:
                  eps.p[ yearid[t] ] +
                  # fixed predictors
                  beta.p[ 1 ] * pac_area[ t ] + # area
                  beta.p[ 2 ] * julian_date[ t ] #julian day 
        # linking both model outputs to observations
        y_obs[ t ] ~ dbern( z[ siteid[t], yearid[t] ] * p[ t ] ) 
        
# Model Validation
        # estimated detections from model:
        yhat[ t ] ~ dbern( z[ siteid[t], yearid[t] ] * p[ t ] ) 

        # Bernoulli likelihood of observations (for logistic regression) # from the data
        lik_yobs[ t ] <- ( ( psi[ siteid[t], yearid[t] ] * p[ t ] )^y_obs[ t ] ) *
              ( ( 1 - psi[ siteid[t], yearid[t] ] * p[ t ] )^( 1 - y_obs[ t ] ) )
        # likelihood of estimated detections:# from the model only
        lik_yhat[ t ] <- ( ( psi[ siteid[t], yearid[t] ]* p[ t ] )^yhat[ t ] ) *
            ( ( 1 - psi[ siteid[t], yearid[t] ] * p[ t ] )^( 1 - yhat[ t ] ) )
      } # t

     } # model close
     
     ", fill = TRUE )

sink() # sends it to text
```

## Run Model 1 
```{r}
# name the model
modelname <- "dm1.txt"

# define parameters monitored:
# only keep those relevant for model comparisons
# jags will estimate all of them, but these are the ones I want to keep track of
params <- c('int.phi', # intercept for persistence model 
            'int.gam', # intercept for colonization model
            'int.p', # intercept for detection
            'beta.phi', # fixed coefficients for persistence
            'beta.gam', # fixed coefficients for colonization model
            'beta.p', # coefs for detection
            'eps.p', # random intercept for year in detection model
            'z', # estimated occupancy state
            'psi',
            'phi', # slope for persistence
            'gam', # slope for colonization
            'p', # slope for detection
            'lik_yobs', # likelihood for each occupancy observation (model validation)
            'lik_yhat', # likelihood for occupancy observations predicted by the model (model validation)
            'yhat') # estimated occurrence from model (model validation)

# create initial values for the model coefficients
zst <- matrix(data = 1, nrow = I, ncol = K ) # tell the model to start at occupied state
# pst <- array(data = 1, dim = c(I,K,J) )
inits <- function(){ list( beta.psi = rnorm( 5 ),
                           beta.p = rnorm( 2 ), 
                           z = zst)}

# give it the data:
str(win.data <- list(y_obs = XJ$y_obs, # observed occupancy for each species
                       K = K , J = J, I = I, T = T, # how many yrs/surveys/sites/rows in the long df
                     
                       # site level predictors:
                       X = as.matrix(X[,1:5]), # X are predictors (only the first 5) 

                       # survey level predictors:
                       siteid = XJ$siteid,
                       yearid = XJ$yearid,
                       surveyid = XJ$surveyid,
                       julian_date = XJ$julian_date,
                       pac_area = XJ$pac_area))                

# call JAGS and summarize posteriors:
m1 <-  autojags( win.data, inits = inits, params, modelname,
                 n.chains = nc, n.thin = nt, n.burnin = nb,
                 iter.increment = 20000, max.iter = 500000, 
                 Rhat.limit = 1.02, # a measure of what you consider "converged"
                 save.all.iter = FALSE, parallel = TRUE )

print(m1, dig = 2)
```
why aren't we interested in psi, why is it z instead? i thought z was just year 1?

## Write and Define Model 2 
Added random year intercept to detection model.
```{r}
sink( "dm2.txt" )
cat( "
     model{
     
      # priors
      # for occupancy model:    
      psi1 ~ dbeta( 4, 4 ) #dunif( 0, 1 ) #alternative prior
      
      # define intercepts for ecological models as mean probs:
      int.phi <- log( mean.phi / ( 1 - mean.phi ) )
      mean.phi ~ dbeta( 4, 4 )  # mean persistence prob
      int.gam <- log( mean.gam / ( 1 - mean.gam ) )
      mean.gam ~ dbeta( 4, 4 )  # mean colonization prob
      
      # for detection
      int.p <- log( mean.p / ( 1 - mean.p ) )
      mean.p ~ dbeta( 4, 4 )  #mean detection prob
      
      # random year intercepts
      for ( k in 1:K ){ 
        eps.p[ k ] ~ dnorm( 0, prec.p ) T(-7, 7)
      } # K
      
      # associated precision of random  intercepts:
      prec.p <- 1 / ( sigma.p * sigma.p )
      sigma.p ~ dt( 0, 2.5, 7 ) T( 0, )
      
      # priors for beta predictors:
      for( q in 1:5 ){ #loop over number of predictors
        beta.phi[ q ] ~ dnorm( 0, 0.1 ) #prior for persistence predictors
      }
      for( q in 1:2 ){ #loop over number of predictors
        beta.p[ q ] ~ dnorm( 0, 0.1 ) #prior for detection predictors
      } #Q
     
      # ecological model for occupancy:
        for( i in 1:I ){  # loop over sites
          # estimating psi[,1]:
          z[ i, 1 ] ~ dbern( psi[ i, 1 ] ) 
          psi[ i, 1 ] <- psi1
     
          for ( k in 2:K ) { #loop over years
            # occupancy state:
            z[ i, k ] ~ dbern( psi[ i, k ] ) 
            # probability of occupancy starts on year 2
            psi[ i, k ] <- z[ i, k-1 ] * phi[ i, k-1 ] + 
                        ( 1 - z[ i, k-1 ] ) * gam[ i, k-1 ] 
              # model for persistence probability:
              logit( phi[ i, k-1 ] ) <- int.phi + 
                  #fixed predictors:
                  beta.phi[ 1 ] * X[ i, 1 ] + # coleogyne
                  beta.phi[ 2 ] * X[ i, 5 ] + # people
                  beta.phi[ 3 ] * X[ i, 4 ] + # geometry
                  beta.phi[ 4 ] * X[ i, 3 ] + # mshrub
                  beta.phi[ 5 ] * X[ i, 2 ] # pjcon
              # model for colonization probability:
              logit( gam[ i, k-1 ] ) <- int.gam 
          } # K years
        } # I sites
        
      # observation model:
      for( t in 1:T ){  #loop over all replicates
        logit( p[ t ] ) <- int.p +
            #random intercept for year
            eps.p[ yearid[t] ] +
            # fixed predictors
            beta.p[ 1 ] * pac_area[ t ] + # area
            beta.p[ 2 ] * julian_date[ t ] # julian day 
        # linking both model outputs to observations
        y_obs[ t ] ~ dbern( z[ siteid[t], yearid[t] ] * p[ t ] ) 
             
        # estimated detections from model:
        yhat[ t ] ~ dbern( z[ siteid[t], yearid[t] ] * p[ t ] ) 
        # Bernoulli likelihood of observations (for logistic regression)
        lik_yobs[ t ] <- ( ( psi[ siteid[t], yearid[t] ] * p[ t ] )^y_obs[ t ] ) *
              ( ( 1 - psi[ siteid[t], yearid[t] ] * p[ t ] )^( 1 - y_obs[ t ] ) )
        # likelihood of estimated detections:
        lik_yhat[ t ] <- ( ( psi[ siteid[t], yearid[t] ]* p[ t ] )^yhat[ t ] ) *
            ( ( 1 - psi[ siteid[t], yearid[t] ] * p[ t ] )^( 1 - yhat[ t ] ) )
      } # t
     } # model close
     
     ", fill = TRUE )

sink()
```

## Run Model 2
```{r}
modelname <- "dm2.txt"

# parameters monitored only keep those relevant for model comparisons
params <- c('int.phi' # intercept for persistence model 
             , 'int.gam' # intercept for colonization model
             , 'int.p' # intercept for detection model
             , 'beta.phi' # fixed coefficients for persistence
             , 'beta.p' # fixed coefficients for detection
             , 'eps.p' # random intercept for year for detection model
             , 'sigma.p' # variance for random intercept
             , 'z' # estimated occupancy state
             , 'psi'
             , 'phi' # estimated persistence
             , 'gam' # estimated colonization
             , 'p' # estimated detection
             , 'lik_yobs' #likelihood for each occupancy observation
             , 'lik_yhat' #likelihood for occupancy observations predicted by the model
             , 'yhat' #estimated occurrence from model
             )

# create initial values for the model coefficients
zst <- matrix(data = 1, nrow = I, ncol = K)

# pst <- array(data = 1, dim = c(I,K,J))

# create initial values to start the algorithm
inits <- function(){list(beta.phi = rnorm(5),
                         beta.p = rnorm(2), 
                         # p = pst,
                         z = zst)
  }

# combine data into object:
str( win.data <- list( y_obs = XJ$y_obs, #observed occupancy for each species
                       K = K , J = J, I = I, T = T
                       # site level predictors
                       ,X = as.matrix(X[,1:5])
                       # survey level predictors
                       ,siteid = XJ$siteid
                       ,yearid = XJ$yearid
                       ,surveyid = XJ$surveyid
                       ,julian_date = XJ$julian_date
                       ,pac_area = XJ$pac_area))                

# call JAGS and summarize posteriors:
m2 <- autojags(win.data, inits = inits, params, modelname,
               n.chains = nc, n.thin = nt, n.burnin = nb,
               iter.increment = 20000, max.iter = 500000, 
               Rhat.limit = 1.02,
               save.all.iter = FALSE, parallel = TRUE) 

print(m2, dig = 2)
```

## Write Model 3
Multi-season occupancy model (different from dynamic because it excludes extinction and colonization parameters).
```{r}
sink( "dm3.txt" )
cat( "
     model{
     
      #priors
      # define intercept for occupancy
      # note that this is using the logit transformation
      
  # occupancy probability
      int.psi <- log( mean.psi / ( 1 - mean.psi ) )
      mean.psi ~ dbeta( 4, 4 )  
      
# it allows us to set the prior on the real scale where the intercept represent the mean probability averaged across sites and years...here we use a beta distribution which ranges between 0 to 1, the parameters given to this prior should result in a tight, bell-shaped distribution centered around 0.5 - plot different distributions to see what they look like.
      
# This same process is used to assign our prior for the intercept in the detection model
      
  # detection probability
      int.p <- log( mean.p / ( 1 - mean.p ) )
      mean.p ~ dbeta( 4, 4 )
     
  # random year intercepts
      for ( k in 1:K ){ 
        eps.k[ k ] ~ dnorm( 0, prec.k ) T(-7, 7) # random year intercept for ecological model
        eps.p[ k ] ~ dnorm( 0, prec.p ) T(-7, 7) # random year intercept for observation model
      } # K
      
  # random site intercept 
      for ( i in 1:I ){
        eps.i[ i ] ~ dnorm( 0, prec.i ) T(-7, 7) # random site intercept for ecological model
       } # i
      
 # error terms for random intercepts for detection model:
      # associated precision of random intercepts for detection is estimated by 1/variance 
      prec.p <- 1 / ( sigma.p * sigma.p )
      sigma.p ~ dt( 0, 2.5, 7 ) T( 0, )
      
 # error terms for the random intercepts for ecological model:
      # random intercept for year
      prec.k <- 1 / ( sigma.k * sigma.k )
      sigma.k ~ dt( 0, 2.5, 7 ) T( 0, )
      
      # random intercept for site
      prec.i <- 1 / ( sigma.i * sigma.i )
      sigma.i ~ dt( 0, 2.5, 7 ) T( 0, )
    
# priors for fixed effects (coefficients) in ecological model:
      for( q in 1:7 ){ # loop over number of predictors
        beta.psi[ q ] ~ dnorm( 0, 0.1 ) # this precision is not fully uninformative but actually regularizes the estimates around 0 to help computation
      }
      
# priors for fixed coefficients in detection submodel:
      for( q in 1:2 ){ # loop over number of predictors
        beta.p[ q ] ~ dnorm( 0, 0.1 ) # prior for detection predictors
      
      } # Q
# END OF PRIORS 
    
# occupancy model:
        for( i in 1:I ){  # loop over sites
          for ( k in 1:K ) { # loop over years
            # true occupancy state, z, is given a bernoulli distribution with probability psi:
            z[ i, k ] ~ dbern( psi[ i, k ] ) 
              
              # probability of occupancy, psi is linked to predictors using a logit function:
              logit( psi[ i, k ] ) <- int.psi + 
                  # random intercepts:
                  eps.k[ k ] + 
                  eps.i[ i ] +
                  
                  # fixed effects:
                  beta.psi[ 1 ] * X[ i, 5 ] + # people
                  beta.psi[ 2 ] * X[ i, 1 ] + # blackbrush
                  beta.psi[ 3 ] * X[ i, 2 ] + # pjcon
                  beta.psi[ 4 ] * X[ i, 3 ] + # mshrub
                  beta.psi[ 5 ] * X[ i, 6 ] + # supai
                  beta.psi[ 6 ] * X[ i, 7 ] + # redwall
                  beta.psi[ 7 ] * X[ i, 4 ]   # canyon geometry
          } # K years
        } # I sites
    

# detection model:
      for( t in 1:T ){  # loop over all replicates
      
        # we link probability of detection, p, to predictors using a logit function
        logit( p[ t ] ) <- int.p +
            # random intercept:
            eps.p[ yearid[t] ] +
      
            # fixed effects:
            beta.p[ 1 ] * pac_area[ t ] + # area
            beta.p[ 2 ] * julian_date[ t ] # julian day 
             
# Here we link our observations to the estimated, true occupancy, z, from our ecological model above
        y_obs[ t ] ~ dbern( z[ siteid[t], yearid[t] ] * p[ t ] ) 
             
             
# For Model Evaluation:     

# Here we estimate  what the model would have produced as observations.
      
        yhat[ t ] ~ dbern( z[ siteid[t], yearid[t] ] * p[ t ] ) 
        
# To finish, we also estimate the likelihood of observed and predicted values
    
    # likelihood of observations
        lik_yobs[ t ] <- ( ( psi[ siteid[t], yearid[t] ] * p[ t ] )^y_obs[ t ] ) *
              ( ( 1 - psi[ siteid[t], yearid[t] ] * p[ t ] )^( 1 - y_obs[ t ] ) )
              
    # likelihood of estimated detections:
        lik_yhat[ t ] <- ( ( psi[ siteid[t], yearid[t] ]* p[ t ] )^yhat[ t ] ) *
            ( ( 1 - psi[ siteid[t], yearid[t] ] * p[ t ] )^( 1 - yhat[ t ] ) )
      } # t
     } # model close
     
     ", fill = TRUE )

sink()
```

## Run Model 3
```{r}
modelname <- "dm3.txt"
# parameters monitored - only keep those relevant for model comparisons

params <- c('int.psi' # intercept for occupancy model
            ,'int.p' # intercept for detection
            ,'beta.psi' # fixed coefficients for occupancy
            ,'beta.p' # fixed coefficients for detection
            ,'eps.k' # random year intercept for occupancy 
            ,'sigma.k' # standard deviation associated with random year for occupancy
            ,'eps.p' # random year intercept for detection 
            ,'sigma.p' # standard deviation associated with random year for detection
            ,'eps.i'
            ,'sigma.i'
            ,'z' # estimated occupancy state
            ,'psi' # probability of occupancy
            ,'p' # probability of detection
            ,'lik_yobs' # likelihood for each occupancy observation
            ,'lik_yhat' # likelihood for occupancy observations predicted by the model
            ,'yhat') # estimated occurrence from model

# create initial values for the model coefficients
zst <- matrix(data = 1, nrow = I, ncol = K )
# pst <- array(data = 1, dim = c(I,K,J) )

# create initial values to start the algorithm
inits <- function(){ list( beta.psi = rnorm( 7 ),
                           beta.p = rnorm( 2 ), 
                           z = zst)
  }
# combine data into object:
str( win.data <- list( y_obs = XJ$y_obs, # observed occupancy for each species
                       K = K , J = J, I = I, T = T,
                       # site level predictors
                       X = as.matrix(X[,1:7]),
                       # survey level predictors
                       siteid = XJ$siteid,
                       yearid = XJ$yearid,
                       surveyid = XJ$surveyid,
                       julian_date = XJ$julian_date,
                       pac_area = XJ$pac_area))                

#call JAGS and summarize posteriors:
m3 <-  autojags( win.data, inits = inits, params, modelname,
                 n.chains = nc, n.thin = nt, n.burnin = nb,
                 iter.increment = 20000, max.iter = 500000, 
                 Rhat.limit = 1.02, # rhat is metric of convergence, it should be verrrrry close to 1
                 save.all.iter = FALSE, parallel = TRUE ) 

print(m3, dig = 2)
# QUESTION: how can I have values greater than 1 here? Isn't this interpretable as occupancy probability and detection probability? 

# posterior predictive plot check for model fit
# pp.check(m3, observed = 'yhat', simulated = 'lik_yhat')
# this is based off of the discrepancy metrics calculated for the real data (for example, the sum of residuals)
```

## Save Model Results
```{r}
# save the workspace
save.image("data/output/bayes_occ_results.RData")

# save the best model
saveRDS(m3, file = "~/Documents/Boise_State/thesis/ms_thesis_ch1/data/output/model_3_results.rds")
```


